---
title: "Online_Retail_Data_Analysis"
author: "Md_Fahd_Al_Georgy"
date: "March 11,2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r loading required packages}
#-----------------------------------------Initial Setup----------------------------------------------
#====================================================================================================
#create a function to check for installed packages and install them if they are not installed
install <- function(packages){
  new.packages <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(new.packages)) 
    install.packages(new.packages, dependencies = TRUE)
  sapply(packages, require, character.only = TRUE)
}

# usage
required.packages <- c("caret", "broom", "ggcorrplot","matrixStats", "tidyverse", "timetk", "prophet", "tidyquant", "modelr", "gridExtra", "grid")
install(required.packages)
options(na.action = na.warn)
```

## Loading and Preprocessing of dataset 

```{r load retail dataset, echo=FALSE}
retail_data <- read_csv("Online Retail.csv",
                   col_types = cols(
                      InvoiceNo = col_character(),
                      StockCode = col_character(),
                      Description = col_character(),
                      Quantity = col_integer(),
                      InvoiceDate = col_datetime("%m/%d/%Y %H:%M"),
                      UnitPrice = col_double(),
                      CustomerID = col_integer(),
                      Country = col_character()
                      )) %>%
  mutate(sell_date = parse_date(format(InvoiceDate, "%Y-%m-%d")),
         day_of_week = wday(sell_date, label = TRUE),
         sell_time = parse_time(format(InvoiceDate, "%H:%M")),
         month = format(InvoiceDate, "%m"),
         earning = Quantity * UnitPrice,
         sell_status = ifelse(Quantity > 0, "sold", "returned"))
```

# Data oraganization


```{r}
write.csv(retail_data, "new_retail_data.csv")
str(retail_data)
```

```{r checking missing value }
colSums(is.na(retail_data))

```
There are missing values in "Description" and "customerID" columns. However, product description and cutomer IDs will not have any impact on our analysis. So, we rather can keep it as it is. 

#Exploratory Data Analysis


```{r unique items }
sapply(retail_data[,c(2,7,8)],function(x)length(unique(x)))
```
We have 4373 customers in our dataset residing across 38 different countries. Form this, we can assume that many among the customers' made repetative transactions throughout the years. 

```{r name of countries  }
unique(retail_data$Country)
```


```{r }
plot1 <- retail_data %>%
  filter(Country == "United Kingdom") %>%
  ggplot(aes(x = Country, fill = sell_status)) +
    geom_bar(alpha = .6) +
    scale_fill_tq(values = palette_dark()) +
    theme_tq() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    guides(fill = FALSE) +
    labs(x = "")

plot2 <- retail_data %>%
  filter(Country != "United Kingdom") %>%
  ggplot(aes(x = Country, fill = sell_status)) +
    geom_bar(alpha = .6) +
    scale_fill_tq(values = palette_green()) +
    theme_tq() +
    theme(legend.position = "right") +
    theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = .9)) +
    labs(x = "",fill = "")
title <- textGrob("Country-wise Selling Trend", gp = gpar(fontface = "bold", cex = 1.5))
grid.arrange(plot1, plot2, top=title ,widths = c(0.2, 1.2))
```

```{r}
retail_data %>%
  ggplot(aes(x = sell_date, color = sell_status)) + ggtitle("Number of sells/returns over time") + 
  theme(plot.title = element_text(hjust = 0.5))+
    facet_grid(sell_status ~ ., scales = "free") +
    geom_freqpoly(bins = 80, size = 1, alpha = 0.9) +
    scale_color_tq(values = palette_dark()) 
```


```{r }
retail_data %>%
  ggplot(aes(x = sell_date, y = ..density.., color = sell_status)) + ggtitle("Relationship between sell and return over time") + 
  theme(plot.title = element_text(hjust = 0.5))+
    geom_freqpoly(size = 1, alpha = 0.9, bins = 60) +
    scale_color_manual(values = palette_light()) 
    
```

# Purchases over the time of the day
```{r}
retail_data %>%
  ggplot(aes(x = sell_time, y = sell_date)) + ggtitle("Transactions trend over the time of the day") +   theme(plot.title = element_text(hjust = 0.5))+
    stat_bin_2d(alpha = 1, bins = 19, color = "grey") +
    scale_fill_gradientn(colours = c(palette_green()[[1]], palette_dark()[[2]])) 
```

# Earnings per day over the months of the year
```{r}
retail_data %>%
  mutate(day = format(InvoiceDate, "%d")) %>%
  group_by(month, day) %>%
  summarise(Earnings = sum(earning)) %>%
  ggplot(aes(x = month, y = day, fill = Earnings)) + ggtitle("Total earnings per day over the months") +   theme(plot.title = element_text(hjust = 0.5))+ ylab("Days of the month")+ 
    geom_tile(alpha = 1, color = "grey") +
    scale_fill_gradientn(colours = c(palette_green()[[1]], palette_dark()[[2]])) +
    theme(legend.position = "right") 
```



```{r}
retail_data %>%
  group_by(StockCode, Description) %>%
  summarise(Total_Quantity = sum(Quantity)) %>%
  arrange(-Total_Quantity) %>% head()
```

```{r}
p1 <- retail_data %>%
  group_by(StockCode, Description) %>%
  summarise(Total_Quantity = sum(Quantity)) %>%
  ggplot(aes(x = Total_Quantity)) +
    geom_density(fill = palette_light()[[1]], alpha = 0.8) +
    theme_tq()

p2 <- retail_data %>%
  group_by(StockCode, Description) %>%
  summarise(Total_Quantity = sum(Quantity)) %>%
  filter(Total_Quantity > 1) %>%
  ggplot(aes(x = Total_Quantity)) +
    geom_density(fill = palette_light()[[1]], alpha = 0.8) +
    theme_tq()

p3 <- retail_data %>%
  group_by(StockCode, Description) %>%
  summarise(Total_Quantity = sum(Quantity)) %>%
  filter(Total_Quantity > 10000) %>%
  ggplot(aes(x = Total_Quantity)) +
    geom_density(fill = palette_light()[[1]], alpha = 0.8) +
    theme_tq()
title1 <- textGrob("Selling trend of product by quantity", gp = gpar(fontface = "bold", cex = 1.5))    
grid.arrange(p1, p2, p3, ncol = 3,top=title1 ,widths = c(0.6, 0.5, 0.7))
```



```{r}
top_sold_items <- retail_data %>%
  group_by(sell_date, StockCode, Description) %>%
  summarise(sum = sum(Quantity)) %>%
  group_by(StockCode, Description)%>%
  summarise(n = n()) %>%
  arrange(-n)

top_items_transactions <- retail_data %>%
  filter(StockCode %in% top_sold_items$StockCode[1:5]) %>%
  group_by(sell_date, StockCode) %>%
  summarise(sum = sum(Quantity)) %>%
  spread(key = StockCode, value = sum)

retail_data %>%
  filter(StockCode %in% top_sold_items$StockCode[1:5]) %>%
  group_by(sell_date, StockCode, Description) %>%
  summarise(sum = sum(Quantity)) %>%
  ggplot(aes(x = sell_date, y = sum)) +
    facet_wrap(~ StockCode, ncol = 1, scales = "free") +
    geom_line(color = palette_light()[[1]], size = 1, alpha = 0.8) +
    theme_tq() +
    labs(Title= "Top 5 Items' Selling Trends", x = "", 
         y = "Quantity of Sold Items")
```

```{r }
retail_data %>%
  filter(StockCode == "85123A") %>%
  group_by(sell_date, sell_status) %>%
  summarise(Most_sold_item = sum(Quantity)) %>%
  ggplot(aes(x = sell_date, y = Most_sold_item, color = sell_status)) + ggtitle("Transaction History of Most Sold Item") +   theme(plot.title = element_text(hjust = 0.5))+ ylab("Item Quantities")+
    facet_wrap(~ sell_status, ncol = 1, scales = "free") +
    geom_line(size = .4, alpha = 1) + 
    scale_color_tq(values = palette_dark()) 
    
```

```{r }
multiple_transactions <- retail_data %>%
  group_by(sell_date, CustomerID) %>%
  summarise(sum = sum(Quantity)) %>%
  group_by(CustomerID) %>%
  summarise(n = n()) %>%
  mutate(revisited_customers = ifelse(n > 1, "reoccurring_customers", "one_time_customers"))

length(which(multiple_transactions$revisited_customers == "reoccurring_customers"))
```


```{r }
revisited_customers_sell_date <- left_join(retail_data, multiple_transactions, by = "CustomerID") %>%
  distinct(sell_date, CustomerID, revisited_customers) %>%
  group_by(sell_date, revisited_customers) %>%
  summarise(n = n()) %>%
  spread(key = revisited_customers, value = n)
multiple_transactions %>%
  group_by(revisited_customers) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = "", y = prop, fill = revisited_customers)) +ggtitle("Ratio of one-time customers & revisiting customers") +   theme(plot.title = element_text(hjust = 0.5))+ ylab("")+ xlab("")+
    geom_bar(stat = "identity", alpha = 1) +
    coord_polar("y", start = 0) +
    scale_fill_tq(values = palette_dark()) 
    
```


```{r }
purchases <- retail_data %>%
  group_by(sell_date, CustomerID) %>%
  summarise(n = n(),
            Tot_item = sum(Quantity),
            Tot_earning = sum(earning)) %>%
  group_by(sell_date) %>%
  summarise(mean_earning_per_customer = mean(Tot_earning),
            mean_quantity_sold_per_customer = mean(Tot_item),
            mean_items_sold_per_customer = mean(n))
purchases %>%
  gather(x, y, mean_earning_per_customer:mean_items_sold_per_customer) %>%
  ggplot(aes(x = sell_date, y = y)) + ggtitle("Selling Trend per Customer") +   theme(plot.title = element_text(hjust = 0.5))+ ylab("")+ xlab("")+
    facet_wrap(~ x, ncol = 1, scales = "free") +
    geom_line(color = palette_light()[[1]], size = 1, alpha = 1) +
    geom_smooth(color = palette_light()[[2]], method = 'loess') 
    
```


```{r}
returned_items <- retail_data %>%
  group_by(sell_date, sell_status) %>%
  summarise(total_quantity = sum(Quantity)) %>%
  spread(key = sell_status, value = total_quantity)
 
returned_items%>%
  gather(x, y, sold:returned) %>%
  ggplot(aes(x = sell_date, y = y, color = x)) + 
    geom_line(size = 1, alpha = 0.8) +
    scale_color_manual(values = palette_light()) +
    theme_tq() +
    labs(x = "", 
         y = "quantity of items",
         title = "Purchase and Return Over the Time")
```

#How many different items are purchased/returned per day?
```{r}
item_transactions <- retail_data %>%
  group_by(sell_date, StockCode) %>%
  summarise(n = n()) %>%
  group_by(sell_date) %>%
  summarise(item_transactions = n())
item_transactions %>%
  ggplot(aes(x = sell_date, y = item_transactions)) +
    geom_line(color = palette_light()[[1]], size = 1, alpha = 0.8) +
    geom_smooth(color = palette_light()[[2]], method = 'loess') +
    theme_tq() +
    labs(title= "Item Transactions Over the Time" ,x = "", 
         y = "number of different items",
         color = "")
```


```{r}
net_earning <- retail_data %>%
  group_by(sell_date) %>%
  summarise(sum_earning = sum(earning),
            mean_earning = mean(earning),
            sum_quantity = sum(Quantity),
            mean_quantity = mean(Quantity))
net_earning %>%
  gather(x, y, sum_earning:mean_quantity) %>%
  ggplot(aes(x = sell_date, y = y)) +
    facet_wrap(~ x, ncol = 1, scales = "free") +
    geom_line(color = palette_light()[[1]], size = 1, alpha = 0.8) +
    geom_smooth(color = palette_light()[[2]], method = 'loess') +
    theme_tq() +
    labs(title="Mean of Earning and Itme Quantity" ,x = "", 
         y = "")
```

# Profit from purchases and returns
```{r }
sold_items <- retail_data %>%
  filter(earning > 0) %>%
  group_by(sell_date) %>%
  summarise(sell_earning = sum(earning),
            #sell_earning_mean = mean(earning),
            sell_quantity = sum(Quantity))
            #sell_quantity_mean = mean(Quantity))
sold_items %>%
  gather(x, y, sell_earning:sell_quantity) %>%
  ggplot(aes(x = sell_date, y = y)) +
    facet_wrap(~ x, ncol = 1, scales = "free") +
    geom_line(color = palette_light()[[1]], size = 1, alpha = 0.8) +
    geom_smooth(color = palette_light()[[2]], method = 'loess') +
    theme_tq() +
    labs(title= "Sell Trends",x = "", 
         y = "")
```


```{r}
return_items <- retail_data %>%
  filter(earning < 0) %>%
  group_by(sell_date) %>%
  summarise(return_items_sum_price = sum(earning),
            mean_return_items_price = mean(earning),
            quantity_return_items = sum(Quantity),
            mean_quantity_return_items = mean(Quantity))
return_items %>%
  gather(x, y, return_items_sum_price:mean_quantity_return_items) %>%
  ggplot(aes(x = sell_date, y = y)) +
    facet_wrap(~ x, ncol = 1, scales = "free") +
    geom_line(color = palette_light()[[1]], size = 1, alpha = 0.8) +
    theme_tq() +
    labs(x = "", 
         y = "")
```


```{r}
unit_quant <- distinct(select(retail_data, sell_date, StockCode, UnitPrice)) %>%
  mutate(unit_quant = paste(sell_date, StockCode, sep = "_")) %>%
  select(unit_quant, UnitPrice)

mean_unit_price <- retail_data %>%
  filter(sell_status == "sold") %>%
  group_by(sell_date, StockCode) %>%
  summarise(n = n()) %>%
  mutate(unit_quant = paste(sell_date, StockCode, sep = "_")) %>%
  left_join(unit_quant, by = "unit_quant") %>%
  group_by(sell_date, StockCode) %>%
  summarise(mean = mean(UnitPrice)) %>%
  group_by(sell_date) %>%
  summarise(mean_unit_price = mean(mean))
mean_unit_price %>%
  ggplot(aes(x = sell_date, y = mean_unit_price)) +
    geom_line(color = palette_light()[[1]], size = 1, alpha = 0.8) +
    theme_tq() +
    labs(x = "", 
         y = "mean unit price of sold items")
```


# Forcasting
```{r }
transaction_per_day <- distinct(select(retail_data, sell_date, day_of_week, month)) %>%
  left_join(net_earning, by = "sell_date") %>%
  left_join(mean_unit_price, by = "sell_date") %>%
  left_join(sold_items, by = "sell_date") %>%
  left_join(return_items, by = "sell_date") %>%
  left_join(purchases, by = "sell_date") %>%
  left_join(revisited_customers_sell_date, by = "sell_date") %>%
  left_join(returned_items, by = "sell_date") %>%
  left_join(item_transactions, by = "sell_date") %>%
  left_join(top_items_transactions, by = "sell_date") %>%
  mutate(diff_sum_earning = sell_earning - lag(sell_earning),
         season = ifelse(month %in% c("03", "04", "05"), "spring",
                         ifelse(month %in% c("06", "07", "08"), "summer",
                                ifelse(month %in% c("09", "10", "11"), "fall", "winter"))))
```

```{r}
transaction_per_day <- transaction_per_day %>%
  mutate(model = ifelse(sell_date <= "2011-10-01", "train", "test"))

colnames(transaction_per_day)[grep("^[0-9]+", colnames(transaction_per_day))] <- paste0("P_", colnames(transaction_per_day)[grep("^[0-9]+", colnames(transaction_per_day))])

```

```{r}
transaction_per_day %>%
  ggplot(aes(x = sell_date, y = sell_earning, color = model)) +
    geom_point(alpha = 0.5) +
    geom_line(alpha = 0.5) +
    scale_color_manual(values = palette_light()) +
    theme_tq()
```

```{r}
augmented_transactions <- transaction_per_day %>%
  rename(date = sell_date) %>%
  select(model, date, sell_earning) %>% 
  tk_augment_timeseries_signature() %>%
  select(-contains("month"))
  
augmented_transactions <- augmented_transactions[complete.cases(augmented_transactions), ]
```

```{r}
(var <- data.frame(colnames = colnames(augmented_transactions[, sapply(augmented_transactions, is.numeric)]),
           colvars = colVars(as.matrix(augmented_transactions[, sapply(augmented_transactions, is.numeric)]))) %>%
  filter(colvars == 0))
```

```{r}
augmented_transactions <- select(augmented_transactions, -one_of(as.character(var$colnames)))
```

#Removing Highly correlated features

```{r}
relation <- cor(augmented_transactions[, sapply(augmented_transactions, is.numeric)])
p.relation <- cor_pmat(augmented_transactions[, sapply(augmented_transactions, is.numeric)])

ggcorrplot(relation,  type = "upper", outline.col = "white", hc.order = TRUE, p.mat = p.relation,
           colors = c(palette_light()[1], "white", palette_light()[2]))
```

```{r}
correlation <- findCorrelation(relation, cutoff=0.8) 
augmented_transactions <- select(augmented_transactions, -one_of(colnames(relation)[correlation]))
train <- filter(augmented_transactions, model == "train") %>%
  select(-model)
test <- filter(augmented_transactions, model == "test")
```

```{r}
fit_lm <- glm(sell_earning ~ ., data = train)
tidy(fit_lm) %>%
  gather(x, y, estimate:p.value) %>%
  ggplot(aes(x = term, y = y, color = x, fill = x)) +
    facet_wrap(~ x, scales = "free", ncol = 2) +
    geom_bar(stat = "identity", alpha = 0.8) +
    scale_color_manual(values = palette_light()) +
    scale_fill_manual(values = palette_light()) +
    theme_tq() + labs(x="", y="")+
    theme(axis.text.x = element_text(angle = 75, vjust = 1, hjust = 1))
```

```{r}
augment(fit_lm) %>%
  ggplot(aes(x = date, y = .resid)) +
    geom_hline(yintercept = 0, color = "red") +
    geom_point(alpha = 0.5, color = palette_light()[[1]]) +
    geom_smooth() +
    theme_tq()
```


#Visualizing the Prediction test
```{r}
pred_test <- test %>%
  add_predictions(fit_lm, "pred_lm") %>%
  add_residuals(fit_lm, ".resid_lm")
```

```{r}
pred_test %>%
    ggplot(aes(x = date, y = .resid_lm)) +
    geom_hline(yintercept = 0, color = "red") +
    geom_point(alpha = 0.5, color = palette_light()[[1]]) +
    geom_smooth() +
    theme_tq()
```


```{r}
pred_test %>%
  gather(x, y, sell_earning, pred_lm) %>%
  ggplot(aes(x = date, y = y, color = x)) +
    geom_point(alpha = 0.5) +
    geom_line(alpha = 0.5) +
    scale_color_manual(values = palette_light()) +
    theme_tq()
```

# Forcasting

```{r}
# Extract index
time_index <- transaction_per_day %>%
    tk_index()

```

```{r }
augmented_transactions %>%
  ggplot(aes(x = date, y = diff)) +
    geom_point(alpha = 0.5, aes(color = as.factor(diff))) +
    geom_line(alpha = 0.5) +
    scale_color_manual(values = palette_light()) +
    theme_tq()
```


```{r }
augmented_transactions %>%
  select(date, wday.lbl, diff) %>%
  filter(wday.lbl != "Sunday" & diff > 86400) %>%
  mutate(days_missing = diff / 86400 -1)
```

```{r}
off_days <- c("2010-12-24", "2010-12-25", "2010-12-26", "2010-12-27", "2010-12-28", "2010-12-29", "2010-12-30", "2010-01-01", "2010-01-02", "2010-01-03",
              "2011-04-22", "2011-04-23", "2011-04-24", "2011-04-25", "2011-05-02", "2011-05-30", "2011-08-29", "2011-04-29", "2011-04-30") %>% 
  ymd()
```

```{r}
future_time_index <- time_index %>%
    tk_make_future_timeseries(n_future = 300, inspect_weekdays = TRUE, inspect_months = FALSE, skip_values = off_days)
future_time_index %>%
    tk_get_timeseries_signature() %>%
    ggplot(aes(x = index, y = diff)) +
    geom_point(alpha = 0.5, aes(color = as.factor(diff))) +
    geom_line(alpha = 0.5) +
    scale_color_manual(values = palette_light()) +
    theme_tq()
```


```{r }
future_data <- future_time_index %>%
    tk_get_timeseries_signature() %>%
    rename(date = index)

prediction <- predict(fit_lm, newdata = future_data)

prediction <- future_data %>%
    select(date) %>%
    add_column(sell_earning = prediction)
transaction_per_day %>%
  select(sell_date, sell_earning) %>%
  rename(date = sell_date) %>%
  rbind(prediction) %>%
  ggplot(aes(x = date, y = sell_earning)) +
    scale_x_date() +
    geom_vline(xintercept = as.numeric(max(transaction_per_day$sell_date)), color = "red", size = 1) +
    geom_point(alpha = 0.5) +
    geom_line(alpha = 0.5) +
    theme_tq()
```

```{r}
residual_testing <- pred_test$.resid_lm
sd_of_res_testing <- sd(residual_testing, na.rm = TRUE)

prediction <- prediction %>%
    mutate(
        lo.95 = sell_earning - 1.96 * sd_of_res_testing,
        lo.80 = sell_earning - 1.28 * sd_of_res_testing,
        hi.80 = sell_earning + 1.28 * sd_of_res_testing,
        hi.95 = sell_earning + 1.96 * sd_of_res_testing
        )
```


```{r}
transaction_per_day %>%
  select(sell_date, sell_earning) %>%
  rename(date = sell_date) %>%
  ggplot(aes(x = date, y = sell_earning)) +
    geom_point(alpha = 0.5) +
    geom_line(alpha = 0.5) +
    geom_ribbon(aes(ymin = lo.95, ymax = hi.95), data = prediction, 
                fill = "#D5DBFF", color = NA, size = 0) +
    geom_ribbon(aes(ymin = lo.80, ymax = hi.80, fill = key), data = prediction,
                fill = "#596DD5", color = NA, size = 0, alpha = 0.8) +
    geom_point(aes(x = date, y = sell_earning), data = prediction,
               alpha = 0.5, color = palette_light()[[2]]) +
    geom_smooth(aes(x = date, y = sell_earning), data = prediction,
                method = 'loess', color = "white") +
    theme_tq()
```

# Prediction using another model

```{r}
transaction_per_day <- transaction_per_day %>%
  mutate(model = ifelse(sell_date <= "2011-10-01", "train2", "test2"))

train2 <- filter(transaction_per_day, model == "train2") %>%
  select(sell_date, sell_earning) %>%
  rename(ds = sell_date,
         y = sell_earning)

test2 <- filter(transaction_per_day, model == "test2") %>%
  select(sell_date, sell_earning) %>%
  rename(ds = sell_date)
```

```{r}
closed <- data.frame(ds = as.Date(c("2010-12-24", "2010-12-25", "2010-12-26", "2010-12-27", "2010-12-28", 
                                      "2010-12-29", "2010-12-30", "2010-01-01", "2010-01-02", "2010-01-03",
                                      "2011-04-22", "2011-04-23", "2011-04-24", "2011-04-25", "2011-05-02", 
                                      "2011-05-30", "2011-08-29", "2011-04-29", "2011-04-30"))) %>%
  mutate(holiday = paste0("off_day_", seq_along(1:length(ds))))
prophet_model_test <- prophet(train2, 
                              growth = "linear", # growth curve trend
                              n.changepoints = 100, # Prophet automatically detects changes in trends by selecting changepoints from the data
                              yearly.seasonality = FALSE, # yearly seasonal component using Fourier series
                              weekly.seasonality = TRUE, # weekly seasonal component using dummy variables
                              holidays = closed) 
```

# Predicting test data
```{r}
forecasting <- predict(prophet_model_test, test2)
forecasting %>%
  mutate(.resid2 = trend - yhat) %>%
  ggplot(aes(x = ds, y = .resid2)) +
    geom_hline(yintercept = 0, color = "red") +
    geom_point(alpha = 0.5, color = palette_light()[[1]]) +
    geom_smooth() +
    theme_tq()
```


```{r}
future <- make_future_dataframe(prophet_model_test, periods = 300)
forecast <- predict(prophet_model_test, future)
plot(prophet_model_test, forecast) +
    theme_tq()


```



